---
title: "Project Bioinformatic Resources"
author: "Annalisa Xamin, Nicola Perotti"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      tidy.opts=list(width.cutoff = 80),
                      tidy = TRUE,
                      fig.align = "center"
                      )

# Set working directory
setwd("~/Documents/Repo_Git/Project_BR")
```


# Introduction
The following analysis focus on RNA-seq count data extracted from different cancer datasets from the Cancer Genome Atlas (TCGA). From the original TCGA data 50 cases (tumor samples) and 50 controls (normal samples) were randomly selected.

# Analysis
The packages needed for the analysis are the following:
```{r load libraries, results='hide'}
library(sessioninfo)
library(tidyverse)
library(biomaRt)
library(MotifDb) # large collection of motifs across different species
library(seqLogo) 
library(PWMEnrich)
library(PWMEnrich.Hsapiens.background)
library(GEOquery)
library(oligo)
library(pd.hg.u133.plus.2)
library(hgu133plus2.db)
library(genefilter)
library(limma)
library(pheatmap)
library(stringr)
library(GenomicFeatures) # to build object with specific information such as genomic coordinates
library(ggplot2)
library(edgeR) # designed to perform Differential Expression Analysis from RNA-Seq data
library(fgsea) # computation of enrichment scores and other statistics
library(org.Hs.eg.db) # database annotation human specific transcript from Ensembl
library(clusterProfiler) # uses Fisher test, explore gene list of interest against a reference
library(enrichplot) # build enrich map
library(ggnewscale)
library(DOSE)
library(pathview) # contains cartoons of pathways and we project on them our genes of interest
```

## Load data
In particular, we consider data coming from **Lung adenocarcinoma**.
```{r Load data}
load("./RData/Lung_adenocarcinoma.RData")
```
After loading the data, the following three data-frames are available:

- `raw_counts_df` which contains the raw RNA-seq counts;

- `c_anno_df`, which contains sample name and condition (case and control);

- `r_anno_df`, which contains the ENSEMBL genes ids, the length of the genes and the genes symbols.

```{r explore raw_counts_df, include=FALSE}
# Explore only the first 5 rows of the dataframe
head(raw_counts_df, n=5)
head(rownames(raw_counts_df), n=5)
head(colnames(raw_counts_df), n=5)
```

```{r explore c_anno_df, include = FALSE}
# Explore only the first 5 rows of the dataframe
head(c_anno_df, n = 5)
head(rownames(c_anno_df), n=5)
head(colnames(c_anno_df)) # there are only 2 columns: sample and condition
```

```{r explore r_anno_df, include = FALSE}
# Explore only the first 5 rows of the dataframe
head(r_anno_df, n = 5)
head(rownames(r_anno_df), n=5)
head(colnames(r_anno_df)) # there are only 3 columns: ensembl_gene_id, external_gene_name, length
```

## Filter protein coding genes
To extract only protein coding genes from `raw_counts_df` and `r_anno_df`, we use the `biomaRt` package.

```{r attributes, include = FALSE}
# get the list of possible attributes
ensembl <- useMart("ensembl")
ensembl <- useDataset("hsapiens_gene_ensembl",mart=ensembl)
attributes <- listAttributes(ensembl)
```


```{r extract protein coding genes}
query <- getBM(attributes=c("ensembl_gene_id",
                            "external_gene_name",
                            "gene_biotype"),
                filters = c("ensembl_gene_id"), 
                values = r_anno_df$ensembl_gene_id,
                mart = ensembl)

query_protein_coding <- query[which(query$gene_biotype == "protein_coding"), ]

```

```{r inspect query_protein_coding, include=FALSE}
query_protein_coding
```

```{r raw_counts r_anno df protein coding}
indexes_r_anno_df <- which(r_anno_df$ensembl_gene_id %in% query_protein_coding$ensembl_gene_id)
r_anno_df_protein_coding <- r_anno_df[indexes_r_anno_df, ]

indexes_raw_counts_df <- which(rownames(raw_counts_df) %in% query_protein_coding$ensembl_gene_id)
raw_counts_df_protein_coding <- raw_counts_df[indexes_raw_counts_df, ]
```

## Differential expression analysis
To perform the differential expression analysis, we will use the `edgeR` package. 

It is important to remove genes with low signal that will have low statistical power. Since, we want to focus on the transcripts we can use to perform the analysis, we can filter raw counts data using a threshold of raw count > 20 in at least 5 replicates.
```{r filtering}
# count threshold
count_thr <- 20

# number of replicates with more counts than the count threshold
repl_thr <- 5

filter_vec <- apply(raw_counts_df_protein_coding,1, # go through all count matrices by rows
    function(y) max(by(y, c_anno_df$condition, function(x) sum(x>=count_thr))))
    # groups the values on each condition and sum all the values above the count
```

```{r see statistics for the filtering, include=FALSE}
table(filter_vec)
```

```{r create filtered df }
filter_counts_df <- raw_counts_df_protein_coding[filter_vec>=repl_thr,]

filter_anno_df <- r_anno_df_protein_coding[rownames(filter_counts_df)]
```


```{r volcano plot}

```

```{r heatmap}

```

## Gene set enrichment analysis
To perform the gene set enrichment analysis, we will use the `clusterProfiler` package.


## Visualization of the enriched pathway


